Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-08-31 Chapter_0/04_Manuscript/Manuscript.tex
Working on paper revision

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[nomarkers, nolists]{endfloat}

@@

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate habitat preferences of dispersers and to create a
permeability surface for the entire KAZA-TFCA. We compared landscape
permeability across different regions within the KAZA-TFCA as well as outside
its boundaries. Lastly, we calculated least-cost paths and corridors to verify
that major movement routes were adequately encompassed within the KAZA-TFCA.

@@

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, observed regional differences in landscape permeability highlight
the need for a coordinated effort towards maintaining or restoring connectivity,
especially where international effort is required.
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become an utmost task for
conservation management \citep{Doerr.2011, Rudnick.2012}, resulting in an
ever-growing number of large and often transboundary protected areas. While
boundaries of such areas are often drawn according to expert opinion and
socio-political needs, subjective assessments have revealed deficiencies in the
past \citep{Clevenger.2002, Pullinger.2010}. Thus, an empirical assessment of
the adequacy of already-existing or planned protected areas using pertinent
animal movement data is paramount for targeted use of valuable and scarce
conservation funds \citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential corridors typically relies on the estimation of permeability surfaces,
which return the ease or willingness at which the focal species traverses a
specific landscape \citep{Sawyer.2011}. Such surfaces are created based on a
species' habitat preferences, which can be quantified using a suite of selection
functions \citep{Zeller.2012}. Specifically, habitat preferences are estimated
by comparing spatial covariates (e.g. environmental and anthropogenic) at
locations visited by the animal to the same spatial covariates at randomly
selected locations \citep{Zeller.2012}. Importantly, selection functions rely on
adequate landscape and relocation data that are representative of the process
being studied \citep{Diniz.2020}. For instance, relocation data collected on
dispersing individuals has been shown to outperform data collected on resident
individuals in the detection of large-scale movement corridors
\citep{Elliot.2014, Diniz.2020}. Nevertheless, dispersal data is inherently
difficult to collect and remains scarce in the connectivity literature
\citep{Vasudev.2015}. As such, most permeability surfaces upon which movement
corridors are identified are created using relocation data collected on resident
individuals. This introduces severe biases and substantially reduces the power
to reveal meaningful movement corridors, for dispersing individuals have
different needs and drives compared to resident individuals \citep{Elliot.2014,
Cozzi.2020}. Such biases have limited our ability to meaningfully assess the
effectiveness of protected areas in securing connectivity for their protected
species.

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at the
appropriate spatial scale \citep{Elliot.2014, Tshipa.2017}. Thus, how well the
boundaries of the KAZA-TFCA reflect natural movement patterns and dispersal
corridors of its most mobile protected species is virtually unknown.

@@

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
analyzed it to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated the degree of selection or avoidance for environmental and
anthropogenic landscape features and used the obtained habitat preferences to
predict a permeability surface spanning the entire KAZA-TFCA. We then
investigated how landscape permeability varies regionally and internationally
and compared permeability within and outside the KAZA-TFCA boundaries. Finally,
we calculated least-cost paths and corridors to identify major movement routes
and to verify that these are successfully covered by the KAZA-TFCA.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on criteria reported in \cite{Behr.2020}, immobilized according
to protocols described in \cite{Osofsky.1996}, and fitted with GPS/Satellite
radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while
still with their natal pack. All procedures were undertaken and supervised by a
Botswana-registered wildlife veterinarian. During dispersal, GPS collars were
programmed to record GPS relocation data every 4 hours and to regularly transmit
them via Iridium satellite system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the Euclidean distance of a relocation to a reference point \citep{Borger.2012},
which in our case was the center of the dispersing coalition's natal home range.
Thus, dispersal was deemed to have started when a coalition had left its natal
home range and continued until the NSD metric remained stationary, implying that
the coalition had successfully settled (Figure S1). In total, we collected 4'169
GPS relocations during dispersal (Figure S2 \& Table S1), resulting in an
average of 260 locations per dispersing coalition (min = 37, max = 729). In our
analysis, we did not differentiate between male and female dispersing
coalitions, for previous research found little differences between sexes during
dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate habitat preferences of dispersing wild dogs, we used a set of
geo-referenced covariates (\Cref{Covariates}) that we aggregated in the
categories \textit{land cover} (which included water cover, distance to water,
shrubs/grassland cover, and tree cover), \textit{protection status} (protected
vs. unprotected), and \textit{human pressure} (which included human influence,
presence of roads, and distance to roads). For each of these covariates we
prepared spatial raster layers from freely available online services or from
remotely sensed satellite imagery. To ensure a consistent resolution (i.e.
cell-size or grain) across covariates, we coarsened or interpolated all layers
to match a resolution of 250m x 250m. For further details on the preparation of
each covariate, see Appendix A.3. We performed processing and manipulation of
data as well as all spatial and statistical analyses using R, version 3.6.1
\citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' selection or avoidance for the above-mentioned spatial
covariates. That is, we paired each realized step (i.e. the connecting line
between two consecutive GPS relocations; \citealp{Turchin.1998}) with 24 random
steps. We generated random steps by sampling turning angles from a uniform
distribution U(\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted using realized steps \citep{Avgar.2016}. A realized step and its 24
associated random steps formed a stratum that received a unique identifier.
Along each step, we extracted the above-mentioned covariates (Table S3),
standardized extracted values using a z-score transformation, and checked for
correlation using Pearson's Correlation Coefficient \(r\). None of the
covariates were overly correlated (\(|r| > 0.6\); \citealp{Latham.2011}) and we
retained all of them for modeling. Our habitat selection model then assumed that
dispersing wild dogs assigned a selection score \(w(x)\) of the following
exponential form to each realized and random step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's preferences for
these covariates (\(\beta_1, \beta_2, ..., \beta_n\)). To estimate habitat
preferences (i.e. the \(\beta\)'s), we used mixed effects conditional logistic
regression analysis as suggested by \cite{Muff.2020}. We implemented their
method using the R-package \textit{glmmTMB} \citep{Mollie.2017} and used
dispersing coalition ID to model random intercepts and slopes. We defined two
movement metrics, namely the cosine of the turning angle (\(cos(ta)\)) and the
log of the step length (\(log(sl)\)), as core covariates and ran stepwise
forward model selection based on Akaike's Information Criterion (AIC;
\citealp{Burnham.2002}) for all other covariates. The inclusion of movement
metrics served to reduce biases in estimated habitat preferences that may have
arisen due to movement preferences \citep{Avgar.2016}. To validate the
predictive power of the most parsimonious habitat selection model, we ran k-fold
cross-validation for case-control studies as described in \cite{Fortin.2009}
(details in Appendix A.5).

@@

\section{Results}
\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)} and
\textit{log(sl)} (\Cref{PermeabilityResults}a). Parameter estimates showed that
dispersing wild dogs moved in a directional and fast manner, as indicated by a
positive selection for small turning angles, i.e. high \(cos(ta)\) (\(\beta\) =
0.14; 95\% CI = 0.07 to 0.21) and longer steps, i.e. high \(log(sl)\) (\(\beta\)
= 0.06, 95\% CI = 0.02 to 0.09). Dispersers avoided moving through water
(\(\beta\) = -0.52, 95\% CI -0.77 to -0.26) but selected for locations in its
vicinity, although the latter effect was not significant (\(\beta\) = -0.32,
95\% CI = -0.72 to 0.08). Dispersers avoided areas that were densely covered by
trees (\(\beta\) = -0.31, CI = -0.46 to -0.15) and preferred areas covered by
shrubs/grassland (\(\beta\) = 0.25, 95\% CI = 0.07 to 0.42). Finally, dispersers
avoided areas that were influenced by humans (\(\beta\) = -0.41, 95\% CI = -0.78
to -0.05).

@@

\begin{figure}[hbtp]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
          \begin{minipage}{\textwidth}
            \includegraphics[width = \textwidth]{99_PermeabilityMap.pdf}
          \end{minipage}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
        \end{scope}
    \end{tikzpicture}
    \caption{Predicted permeability surface for the extent of the KAZA-TFCA.
    Permeability was predicted by calculating selection scores \(w(x) =
    exp(\beta_1 x_1 + \beta_2 x_2 + ... + \beta_n x_n)\) for each raster cell
    based on the raster cell's underlying covariates (\(x_i\)) and estimated
    habitat preferences (\(\beta_i\)). Areas that dispersers find easy to
    traverse are depicted in bright colors. Bold white lines delineate the
    borders of the KAZA-TFCA, whereas dashed white lines show country borders.}
    \label{PermeabilityMap}
  \end{center}
\end{figure}

@@

Our results emphasize that human influences constitute some of the main barriers
to connectivity among wild dog populations. This conforms to findings on
dispersing wild dogs from eastern Africa \citep{Masenga.2016, Oneill.2020} but
conflicts with findings from South Africa by \cite{DaviesMostert.2012}, who
reported a high willingness of dispersers to cross human-dominated landscapes.
We believe that such differences are due to the unavailability of alternative
routes through natural landscapes, which may have forced dispersers in South
Africa to cross human dominated landscapes despite a strong aversion to do so.
In this regard, our representation of dispersal corridors and the resulting
connectivity appear conservative, as dispersers may be able to make the best out
of a bad situation and cross landscapes characterized by considerably
unfavorable conditions \citep{Palomares.2000, Elliot.2014}. Nevertheless,
successful conservation of this species relies on policymakers' and local
authorities' willingness and ability to provide and conserve natural areas that
remain free from anthropogenic pressures. This is not only paramount in light of
increasing connectivity and facilitating dispersal, but also in terms of
reducing human-caused mortality during dispersal. In fact, previous studies have
shown that human-caused mortality represents a major threat to wild dogs'
ability to disperse \citep{Woodroffe.2019, Cozzi.2020}.

@@

Although dispersers avoided moving through water, they selected locations in its
vicinity. This preference may be caused by the occurence of prey close to water
\citep{Bonyongo.2005}. For the same reason, however, competitors such as lions,
spotted hyenas, and resident wild dogs may also use areas close to water
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs to move
into prey-poorer areas away from water. Given the influence that resident
conspecifics, competitors, and prey can have on dispersers \citep{Cozzi.2018,
Armansin.2019} future studies should strive to collect and incorporate intra-
and interspecific relationships into analyses of landscape connectivity.

@@

Our approach of identifying movement corridors based on pre-defined start and
end points implicitly assumes that individuals know the end point of their
dispersal journey and that they have almost complete knowledge of associated
movement costs \citep{Panzacchi.2016}. Since dispersers often move into unknown
territory, this may not necessarily be the case \citep{Abrahms.2017,
Cozzi.2020}. However, specification of pre-defined end points might not be
necessary, as the parametrized iSSF model can be used as mechanistic movement
model to simulate dispersal events from known source points, yet without
restricting the domain of potential end points \citep{Signer.2017}.
Consequently, movement corridors would emerge more naturally as the result of a
myriad of simulated dispersal events. While a simulation-based approach is
conceptually straightforward, it requires a comprehensive mechanistic
understanding of dispersal movements, which is conditional on our ability to
collect additional dispersal data and adequately represent the landscape through
which individuals move.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{todonotes}

@@

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate relative selection strenghts\todo{as per reviewer
1} of dispersers and to create a permeability surface for the entire KAZA-TFCA.
We compared landscape permeability across different regions within the KAZA-TFCA
as well as outside its boundaries. Lastly, we calculated least-cost paths and
corridors to verify that major movement routes were adequately encompassed
within the KAZA-TFCA.

@@

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, observed regional differences in landscape permeability highlight
the need for a coordinated effort towards maintaining or restoring connectivity,
especially where transboundary dispersal occurs.\todo{According to suggestions
by reviewer 2}
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become a task of utmost
\todo{Reviewer one pointed out an awkward use of the word utmost. I have changed
it according to his suggestion.} importance for conservation management
\citep{Doerr.2011, Rudnick.2012}, resulting in an ever-growing number of large
and often transboundary protected areas. While boundaries of such areas are
often drawn according to expert opinion and socio-political needs, subjective
assessments have revealed deficiencies in the past \citep{Clevenger.2002,
Pullinger.2010}. Thus, an empirical assessment of the adequacy of
already-existing or planned protected areas using pertinent animal movement data
is paramount for targeted use of valuable and scarce conservation funds
\citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential corridors typically relies on the estimation of permeability surfaces,
which return the ease or willingness at which the focal species traverses a
specific landscape \citep{Sawyer.2011}. Such surfaces are created based on a
species' relative selection strengths\todo{reviewer 1}, which can be quantified
using a suite of selection functions \citep{Zeller.2012}. Specifically, relative
selection strenghts \todo{reviewer 1} are estimated by comparing spatial
covariates (e.g. environmental and anthropogenic) at locations visited by the
animal to the same spatial covariates at locations avilable to, yet unused by
\todo{Reviewer 1 pointed out correctly that we should talk about available
locations, rather than randomly selected locations} the animal
\citep{Zeller.2012}. Importantly, selection functions rely on adequate landscape
and relocation data that are representative of the process being studied
\citep{Diniz.2020}. For instance, relocation data collected on dispersing
individuals has been shown to outperform data collected on resident individuals
in the detection of large-scale movement corridors \citep{Elliot.2014,
Diniz.2020}. Nevertheless, dispersal data is inherently difficult to collect and
remains scarce in the connectivity literature \citep{Vasudev.2015}. As such,
most permeability surfaces upon which movement corridors are identified are
created using relocation data collected on resident individuals. While some
species use the same habitats during residence and dispersal
\citep{Fattebert.2015}, others depict vastly different habitat selection
depending on the movement mode \citep{Elliot.2014, Abrahms.2017}. Thus, the use
of data collected during residency can introduce severe biases in estimated
landscape permeability and has likely limited our ability to meaningfully assess
the effectiveness of protected areas in securing connectivity for their
protected species.\todo{Reviewer 2 desired to change from: this introduces - to:
this can introduce}

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at large
spatial scales \citep{Elliot.2014, Tshipa.2017}. Thus, how well the boundaries
of the KAZA-TFCA reflect natural movement patterns and dispersal corridors of
its most mobile protected species is virtually unknown.

@@

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
analyzed it to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated the relative selection strengths\todo{reviewer 1} towards or agains
environmental and anthropogenic landscape features and used the obtained habitat
coefficients to predict a permeability surface spanning the entire KAZA-TFCA. We
then investigated how landscape permeability varies regionally and
internationally and compared permeability within and outside the KAZA-TFCA
boundaries. Finally, we calculated least-cost paths and corridors to identify
major movement routes and to verify that these are successfully covered by the
KAZA-TFCA.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on age, number of same‐sex siblings, pack size, and presence of
unrelated individuals of the opposite sex in ther pack citep \citep{McNutt.1996,
Behr.2020}\todo{We originally only pointed to Behr and did not include the
criteria. The editor desired to include those}. Individuals were immobilized
according to protocols described in \cite{Osofsky.1996}, and fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin, Germany}) while still with their natal pack. All procedures were
undertaken and supervised by a Botswana-registered wildlife veterinarian. Fully
assembled collars produced a tag weight of approximately xx kg, which conforms
to the proposed threshold of 3\% of an animal's body weight
\citep{Kenward.2000}. Collars were mounted using leather belts and included a
drop off mechanism triggered by a slowly decomposing piece of cloth.\todo{Editor
desired more information on GPS collar} During dispersal, GPS collars were
programmed to record GPS relocation data every 4 hours and to regularly transmit
them via Iridium satellite system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the squared \todo{Reviewer 1 pointed out correctly that we need to talk about
the \textit{squared} euclidean distance} Euclidean distance of a relocation to a
reference point \citep{Borger.2012}, which in our case was the center of the
dispersing coalition's natal home range. Thus, dispersal was deemed to have
started when a coalition had left its natal home range and continued until the
NSD metric remained stationary, implying that the coalition had successfully
settled (Figure S1). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate habitat relative selection strength\todo{As per reviewer 1} of
dispersing wild dogs, we used a set of geo-referenced covariates
(\Cref{Covariates}) that we aggregated in the categories \textit{land cover}
(which included water cover, distance to water, shrubs/grassland cover, and tree
cover), \textit{protection status} (protected vs. unprotected), and
\textit{human pressure} (which included human density, presence of roads, and
distance to roads). For each of these covariates we prepared spatial raster
layers from freely available online services or from remotely sensed satellite
imagery. To ensure a consistent resolution (i.e. cell-size or grain) across
covariates, we coarsened or interpolated all layers to match a resolution of
250m x 250m. For further details on the preparation of each covariate, see
Appendix A.3. We performed processing and manipulation of data as well as all
spatial and statistical analyses using R, version 3.6.1 \citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' relative selection strentgh \todo{Reviewer 1 suggested
to use relative selection strength, rather than preference} for the
above-mentioned spatial covariates. That is, we paired each realized step (i.e.
the connecting line between two consecutive GPS relocations;
\citealp{Turchin.1998}) with 24 random steps. We generated random steps by
sampling turning angles from a uniform distribution U(\(-\pi, +\pi\)) and step
lengths from a gamma distribution that was fitted using realized steps
\citep{Avgar.2016}. A realized step and its 24 associated random steps formed a
stratum that received a unique identifier. Along each step, we extracted the
above-mentioned covariates (Table S3), standardized extracted values using a
z-score transformation, and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection model then assumed that dispersing wild dogs assigned a selection
score \(w(x)\) of the following exponential form to each realized and random
step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's relative
selection strength \todo{As per reviewer 1} for these covariates (\(\beta_1,
\beta_2, ..., \beta_n\)). To estimate relative selection strengths \todo{As per
reviewer 1} (i.e. the \(\beta\)'s), we used mixed effects conditional logistic
regression analysis as suggested by \cite{Muff.2020}. We implemented their
method using the R-package \textit{glmmTMB} \citep{Mollie.2017} and used
dispersing coalition ID to model random slopes. We also modeled random
intercepts with an arbitrary high variance of 1e6 to make use of the
\textit{poisson trick} (see \cite{Muff.2020}). We defined three movement
metrics, namely the cosine of the turning angle (\(cos(ta)\)), the step length
\(sl_\) and the natural logarithm of the step length (\(log(sl)\)), as core
covariates and ran stepwise forward model selection based on Akaike's
Information Criterion (AIC; \citealp{Burnham.2002}) for all other
covariates.\todo{Reviewer 1 asked for more details in the random effects. I
therefore decided to update this section too} The inclusion of movement metrics
served to reduce biases in estimated habitat selection coefficients \todo{as per
revierwer 1} that may have arisen due to movement behavior \citep{Avgar.2016}.
To validate the predictive power of the most parsimonious habitat selection
model, we ran k-fold cross-validation for case-control studies as described in
\cite{Fortin.2009} (details in Appendix A.5).

@@

\section{Results}
\subsection{Dispersal Events}
In total, we collected 4'169 GPS relocations during dispersal (Figure S2 \&
Table S1), resulting in an average of 261 (\(SD\) = 207) locations per
dispersing coalition. Dispersers on dispersed on average for 48 days (\(SD\) =
44), covered a mean euclidean distance of 54 km (\(SD\) = 71) and a cumulative
distance of 597 km (\(SD\) = 508). One female coalition dispersed far east into
the Hwange Hwange National Park and covered a cumulative distance of over 360 km
in under 9 days.\todo{Editor desired a quick summary of the dispersal data.}

\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)} and
\textit{log(sl)} (\Cref{PermeabilityResults}a). The positive coefficient for
\(log(sl)\) (\(\beta\) = 0.06, 95\% CI = 0.02 to 0.09) showed that our fitted
(yet selection-biased) gamma distribution produced step-lengths slightly shorter
than under a selection-free gamma distribution. Similarly, the positive
coefficient for \(cos(ta)\) (\(\beta\) = 0.14; 95\% CI = 0.07 to 0.21) indicates
that the uniform distribution for turning angles produced turning angles with
too little directionality.\todo{Reviewer 1 nicely explained that the
coefficients for log(sl) and cos(ta) imply that the initial distributions were
selection-biased} With respect to environmental covariates, dispersers avoided
moving through water (\(\beta\) = -0.52, 95\% CI -0.77 to -0.26) but selected
for locations in its vicinity, although the latter effect was not significant
(\(\beta\) = -0.32, 95\% CI = -0.72 to 0.08). Dispersers avoided areas that were
densely covered by trees (\(\beta\) = -0.31, CI = -0.46 to -0.15) and preferred
areas covered by shrubs/grassland (\(\beta\) = 0.25, 95\% CI = 0.07 to 0.42).
Finally, dispersers avoided areas that were influenced by humans (\(\beta\) =
-0.41, 95\% CI = -0.78 to -0.05).

@@

\begin{figure}[hbtp]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
          \begin{minipage}{\textwidth}
            \includegraphics[width = \textwidth]{99_PermeabilityMap.pdf}
          \end{minipage}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
        \end{scope}
    \end{tikzpicture}
    \caption{Predicted permeability surface for the extent of the KAZA-TFCA.
    Permeability was predicted by calculating selection scores \(w(x) =
    exp(\beta_1 x_1 + \beta_2 x_2 + ... + \beta_n x_n)\) for each raster cell
    based on the raster cell's underlying covariates (\(x_i\)) and estimated
    selection strenghts (\(\beta_i\)). Areas that dispersers find easy to
    traverse are depicted in bright colors. Bold white lines delineate the
    borders of the KAZA-TFCA, whereas dashed white lines show country borders.}
    \label{PermeabilityMap}
  \end{center}
\end{figure}

@@

Our results emphasize that human influences constitute some of the main barriers
to connectivity among wild dog populations. According to our model, dispersers
avoided corssing human dominated landscapes whenever given the
choice.\todo{According to reviewer 1 we should point out that our model compares
available vs. unavailable} This conforms to findings on dispersing wild dogs
from eastern Africa \citep{Masenga.2016, Oneill.2020} but conflicts with
findings from South Africa by \cite{DaviesMostert.2012}, who reported a high
willingness of dispersers to cross human-dominated landscapes. We believe that
such differences are due to the unavailability of alternative routes through
natural landscapes, which may have forced dispersers in South Africa to cross
human dominated landscapes despite a strong aversion to do so. In this regard,
our representation of dispersal corridors and the resulting connectivity appear
conservative, as dispersers may be able to make the best out of a bad situation
and cross landscapes characterized by considerably unfavorable conditions
\citep{Palomares.2000, Elliot.2014}. Nevertheless, successful conservation of
this species relies on policymakers' and local authorities' willingness and
ability to provide and conserve natural areas that remain free from
anthropogenic pressures. This is not only paramount in light of increasing
connectivity and facilitating dispersal, but also in terms of reducing
human-caused mortality during dispersal. In fact, previous studies have shown
that human-caused mortality represents a major threat to wild dogs' ability to
disperse \citep{Woodroffe.2019, Cozzi.2020}.

@@

Although dispersers avoided moving through water, they selected locations in its
vicinity. This behavior may be caused by the occurence of prey close to water
\citep{Bonyongo.2005}. For the same reason, however, competitors such as lions,
spotted hyenas, and resident wild dogs may also use areas close to water
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs to move
into prey-poorer areas away from water. This could also explain the effect of
distane to water was insignificant.\todo{Reviewer 1 suggested to add this} Given
the influence that resident conspecifics, competitors, and prey can have on
dispersers \citep{Cozzi.2018, Armansin.2019} future studies should strive to
collect and incorporate intra- and interspecific relationships into analyses of
landscape connectivity.

Overall, our findings on habitat selection of dispersing wild dogs coincide with
findings from Kenya by \cite{Oneill.2020}, suggesting that there are strong
commonalities across ecosystems. Hence, despite wild dogs' ability to cope with
and adapt to a wide range of habitats, we feel that that the fundamental factors
included in our study and their influences can be generalized to other
populations.\todo{Editor desired that we discuss the applicability of our
ecosystem to other ecosystems.}

@@

Our approach of identifying movement corridors based on pre-defined start and
end points implicitly assumes that individuals know the end point of their
dispersal journey and that they have almost complete knowledge of associated
movement costs \citep{Koen.2014, Panzacchi.2016}. Since dispersers often move
into unknown territory, this may not necessarily be the case
\citep{Abrahms.2017, Cozzi.2020}. However, specification of pre-defined end
points might not be necessary, as the parametrized iSSF model can be used as
mechanistic movement model to simulate dispersal events from known source
points, yet without restricting the domain of potential end points
\citep{Signer.2017}. Consequently, movement corridors (i.e. transient
utilisation distribution) would emerge more naturally as the result of a myriad
of simulated dispersal events. A very similar approach could be applied to
produce permeability surfaces from steady state utilistation distributions
\citep{Avgar.2016, Signer.2017}. \citep{Signer.2017} have shown that such
permeability surfaces are less prone to overestimateing permeability in areas
that lie far from suitable habitats. While a simulation-based approach is
conceptually straightforward, computational requirements are tremendous,
especially for such a large extent as constituted by the KAZA-TFCA.
\todo{Reviewer 1 pointed out that the permeability surface likely overestimates
permeability in areas far from suitable habitat. Additionally, reviewer 2
mentioned that our dataset by far suffices for simulations and we should not
argue with sample size.}

Our work shows how dispersal data of a highly mobile species can be used to
\todo{to instead of t, pointed out by the editor and reviewer 1} identify
movement corridors and to assess the adequacy of protected areas. In our case,
the predicted movement corridors of African wild dogs were well contained within
the boundaries of the world's largest transboundary conservation area, namely
the KAZA-TFCA, suggesting that it will significantly contribute to the long-term
viability of this species. Moreover, our connectivity network allowed revealing
potential dispersal hubs through which dispersers gain access to more remote
regions of the study area. Finally, our investigations showed that human
influence constitutes one of the main barriers to dispersal and substantially
reduces landscape connectivity. Successful conservation of wide-ranging species,
such as exemplified by the African wild dog, will therefore be contingent on the
willingness of local authorities, policymakers, and land managers to preserve
areas that remain free from human strains. Ultimately, our work contributes to
the growing field of connectivity studies and provides and easily expandable
framework for assessing the adequacy of already-existing or planned protected
areas.
